---
title: "Statistical Analysis, MSCA 31007, Lecture 2"
author: "Hope Foster-Reyes"
date: "October 7, 2016"
output: pdf_document
---

# Simulation of Random Variables

_Notes_

* _Style Guide: https://google.github.io/styleguide/Rguide.xml#functiondefinition _

* _Packages required: random, compositions._

## Part 1. Generate uniformly distributed random numbers

## 1.1 Use runif().

```{r q1.1}
# Simulate n pseudo-random numbers uniformly distributed on [a,b]
set.seed(15)
sample.runif <- runif(1000, 0, 1)
str(sample.runif)
head(sample.runif)
quantile(sample.runif)
hist(sample.runif, main = "1000 Pseudo-Random Numbers via Runif()")
boxplot(sample.runif, main = "1000 Pseudo-Random Numbers via Runif()")
```

## 1.2 Simulate Uniform Random Sample on [0,1] Using Random.org.

```{r q1.2}
# Load random package to interface with random.org
library(random)

# Download binary sequence from random.org
num.flips <- 1000
data.random.org <- randomNumbers(n = num.flips, min = 0, max = 1, 
                                 col = 1, base = 2, check = TRUE)
head(data.random.org)
```

**Let us turn our sequence of {0,1} into uniform random numbers on [0,1].**

```{r q1.2-transform}
# Load compositions package
suppressMessages(library(compositions))

# Create function that interprets a sequence of zeros and ones of length n as a binary
#   number and converts that binary into decimal
BinaryToDec <- function(binary.seq){
  # Argument binary.seq: vector containing a binary sequence of 0's and 1's
  unbinary(paste(binary.seq, collapse = ""))
}

# Demonstrate function
BinaryToDec(c(1,1,1,1,1,0))

# Turn the sequence of 0's and 1's from random.org into a 100x10 matrix
matrix.binary <- matrix(data.random.org, ncol = 10)
head(matrix.binary)
str(matrix.binary)

# Transform each row into decimal format
vector.decimal <- apply(matrix.binary, 1, BinaryToDec)
str(vector.decimal)

# Divide the numbers by 2^10 to make real numbers in [0,1]
sample.random.org <- vector.decimal / 2^10
sample.random.org
quantile(sample.random.org)
hist(sample.random.org, main = "100 Pseudo-Random Numbers, Transformed from Random.org Binary Data")
boxplot(sample.random.org, main = "100 Pseudo-Random Numbers, Transformed from Random.org Binary Data")
```

All numbers in sample.random.org are between 0 and 1. This is our equivalent of the sample obtained by runif().

## Part 2. Test random number generator

## 2.1 Test uniformity of distribution of both random number generators

### 2.1.1. Sample obtained by runif()

Analyze what was simulated.


